Non-Vertical Cultural Transmission, Assortment, \\and the Evolution of Cooperation

### 2020-11-23 ms/main.tex
mwf comments

% Abstract
\begin{abstract}
Cultural evolution of cooperation under vertical and non-vertical cultural transmission is studied, and conditions are found for fixation and coexistence of cooperation and defection. 
The evolution of cooperation is facilitated by horizontal transmission and by an association between social interactions and horizontal transmission.
The effect of oblique transmission depends on the horizontal transmission bias.
Stable polymorphism of cooperation and defection can occur, and
when it does, reduced association between social interactions and horizontal transmission evolves, which leads to a decreased frequency of cooperation and lower population mean fitness.
The deterministic conditions are compared to outcomes of stochastic simulations of structured populations.
Parallels are drawn with Hamilton's rule incorporating assortment and effective relatedness.
\end{abstract}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%
Cooperation and defection can coexist at frequencies $\hat{p}^*$ and $1-\hat{p}^*$ (\autoref{eq:general_equilibrium}). 
When it is feasible, this equilibrium is stable or unstable under the conditions of Result \ref{result:vert_obli_hori}, parts 3 and 4, respectively. The yellow and blue areas in Figures~\ref{fig:equilibria_v1} and \ref{fig:equilibria} show cases of stable and unstable polymorphism, respectively.
When $\hat{p}^*$ is unstable, cooperation will fix if its initial frequency is $\hat{p}>\hat{p}^*$, and defection will fix if $\hat{p}<\hat{p}^*$.
$\hat{p}^*$ is unstable when there is horizontal transmission bias for cooperation, $T_A>T_B$, and the cost is intermediate, $\gamma_1 < c < \gamma_2$.
\autoref{fig:equilibria_v1}d shows the mapping $\hat{p} \to \hat{p}'-\hat{p}$.

@@

\begin{figure}[p]
  \centering       
    \includegraphics[width=0.45\textwidth]{Result3_TA_TB.eps}~
    \includegraphics[width=0.45\textwidth]{Result3_TB_TA.eps}
    \includegraphics[width=0.45\textwidth]{Result3_alpha_Vs_v_TA_TB.eps}~
    \includegraphics[width=0.45\textwidth]{Result3_alpha_Vs_v_TB_TA.eps}
    \caption{\textbf{Evolution of cooperation under vertical, oblique, and horizontal cultural transmission.} 
    The figure shows parameter ranges for global fixation of cooperation (green), global fixation of defection (red), fixation of either cooperation or defection depending on the initial conditions, i.e. unstable polymorphism (blue), and stable polymorphism of cooperation and defection (yellow).
	In all cases the vertical transmission rate $v$ is on the x-axis.
	(\textbf{a-b})~Cost of cooperation $c$ is on the y-axis and the cost thresholds $\gamma_1$ and $\gamma_2$ (\autoref{eq:cost_thresholds_v_threshold}) are represented by the solid and dashed lines, respectively. 
    (\textbf{c-d})~interaction-transmission association $\alpha$ is on the y-axis and the interaction-transmission association thresholds $a_1$ and $a_2$ (\autoref{eq:boundries_assortative_meeting_general_case}) are represented by the solid and dashed lines, respectively. 
    Horizontal transmission is biased in favor of cooperation, $T_A>T_B$, in (\textbf{a}) and (\textbf{c}), or defection, $T_A<T_B$, in (\textbf{b}) and (\textbf{d}).    
    Here, $T_A = 0.5$, and
    (\textbf{a}) $b=1.2$, , $T_B = 0.4$, $\alpha = 0.4$;
    (\textbf{b}) $b=2$, $T_B = 0.7$, $\alpha = 0.7$;
    (\textbf{c}) $b=1.2$, $T_B = 0.4$, $c=0.5$;
    (\textbf{d}) $b=2$, $T_B = 0.7$, $c=0.5$.
    }
    \label{fig:equilibria}
\end{figure}

@@

\begin{figure}[p]
  \centering       
    \includegraphics[width=0.45\textwidth]{Result2_c_zero_alpha.eps}
	  ~
    \includegraphics[width=0.45\textwidth]{Result2_c_non_zero_alpha.eps}
    ~
    \includegraphics[width=0.45\textwidth]{Result2_alpha.eps}   
    ~
    \includegraphics[width=0.48\textwidth]{coexistence_without_oblique.eps}
    \caption{\textbf{Evolution of cooperation under vertical and horizontal cultural transmission (v=1).} 
    The figure shows parameter ranges for global fixation of cooperation (green), global fixation of defection (red), fixation of either cooperation or defection depending on the initial conditions, i.e. unstable polymorphism (blue), and stable polymorphism of cooperation and defection (yellow).
    (\textbf{a-c})~The horizontal transmission bias ($T_A-T_B$) is on the x-axis.
    In panels (a) and (b), the cost of cooperation $c$ is on the y-axis and the cost thresholds $\gamma_1$ and $\gamma_2$ (\autoref{eq:cost_thresholds_v_threshold}) are the solid and dashed lines, respectively. 
	In panel (c), interaction-transmission association $\alpha$ is on the y-axis and the interaction-transmission association thresholds $a_1$ and $a_2$ (\autoref{eq:boundries_assortative_meeting_general_case}) are the solid and dashed lines, respectively.
    Here, $b=1.3$, $T_A=0.4$, $v=1$, (a)~$\alpha = 0$, (b)~$\alpha = 0.7$, (c)~$c = 0.35$.    
    \textbf{(d)} Change in frequency of cooperation among juveniles ($\hat{p}'-\hat{p}$) as a function of the frequency ($\hat{p}$), see \autoref{eq:nextgen_juveniles}.
    The orange curve shows convergence to a stable polymorphism ($T_A = 0.4$, $T_B = 0.9$, $b = 12$, $c=0.35$, $v=1$, and $\alpha = 0.45$).
    The blue curve shows fixation of either cooperation or defection, depending on the initial frequency ($T_A = 0.5$, $T_B = 0.1$, $b = 1.3$, $c=0.904$, $v=1$, and $\alpha = 0.4$).
    Markers show the three equilibria.
  	}
    \label{fig:equilibria_v1}
\end{figure}

@@

%%% Figure: Spatial model - global selection
\begin{figure}[p]
  \centering
	\includegraphics[width=0.3\textwidth]{{Ta=Tb_b=1.3_globalSelection_a}.pdf}
    ~
	\includegraphics[width=0.3\textwidth]{{Ta=0.4_Tb=0.3-0.5_b=1.3_globalSelection_withSmallBox_b}.pdf} 
    ~
    \includegraphics[width=0.36\textwidth]{{timeSereis_b=1.3_TA=0.4_TB=0.435_c=0.02_globalSelection_normalizeByNumInteractions_c}.pdf}
  \caption{
  \textbf{Evolution of cooperation in a structured population.}
	\textbf{(a-b)}~The expected frequency of cooperators in a structured population after 10,000 generations is shown (red for 0\%, green for 100\%) as a function of both the cost of cooperation, $c$, on the y-axis, and either the symmetric horizontal transmission rate, $T=T_A=T_B$, on the x-axis of panel~(a), or the transmission bias, $T_A-T_B$, on the x-axis of panel~(b).
  Black curves represent the cost thresholds for the evolution of cooperation in a well-mixed population with interaction-transmission association, where $\alpha=1/8$ in \autoref{eq:equal_transmission} for  panel~(a) and in \autoref{eq:cost_thresholds_v_threshold} for~(b).
	The inset in panel~(b) focuses on an area of the parameter range in which neither phenotype is fixed throughout the simulation, maintaining a stochastic locally stable polymorphism~\citep{Karlin1975}.
	This stochastic polymorphism is illustrated in panel (c), which shows the frequency of cooperators (green) and defectors (red) over time for the parameter set marked by an~\textit{x} in panel~(b).
  In all cases, the population evolves on a 100-by-100 grid.
  Cooperation and horizontal transmission are both local between neighboring sites, and each site has 8 neighbors.
  Selection operates globally (see \autoref{fig:spatial_local_selection} for results from a model with local selection).
  Simulations were stopped at generation 10,000 or if one of the phenotypes fixed. 50 simulations were executed for each parameter set.
  Benefit of cooperation, $b=1.3$; perfect vertical transmission $v=1$;
  \textbf{(a)} Symmetric horizontal transmission, $T=T_A=T_B$;
  \textbf{(b)} Horizontal transmission rates $T_A=0.4$, $0.3<T_B<0.5$;
  \textbf{(c)} Horizontal transmission rates $T_A = 0.4 < T_B = 0.435$ and cost of cooperation $c=0.02$.
  }
  \label{fig:spatial}
\end{figure}

@@

%%% Figure: frequency dynamics
\begin{figure}[p]
  \centering
    \includegraphics[scale=0.45]{Time_Figure_Equal_Horizontal.pdf}
    ~\\
    \includegraphics[scale=0.45]{Time_Figure_Only_Vertical_No_Alpha.pdf}
    ~\\
    \includegraphics[scale=0.45]{Time_Figure_No_Vertical.pdf}
  \caption{
  \textbf{Dynamics of the frequency of cooperation.}
  The frequency $\tilde{p}$ of cooperator parents in \textbf{(a-b)} and the frequency $p$ of adults cooperators in \textbf{(c)}.
  The different lines correspond to parameter values that lead to fixation of cooperation (green), extinction of cooperation (red), or stable polymorphism of cooperators and defectors (yellow).
  \textbf{(a)} $v=1$, $T_A=T_B=T = 0.2$, $\alpha = 0.5 \neq 0$, $\tilde{p}_0 = 0.5$ and $c=0.1$; \textbf{(b)} $v=1$, $\alpha = 0$, $\tilde{p}^* \approx 0.35$, $T_A = 0.65$, $T_B = 0.1$, $b=1.3$ and $c=0.65$; \textbf{(c)} $v=0$, $\alpha =0.5$, $p_0 = 0.5$, $T_A = 0.5$, $b=1.3$ and $c = 0.5$.
  }
  \label{fig:results}
\end{figure}

%%% Figure: Spatial model - local selection
\begin{figure}[h]
  \centering
   \includegraphics[scale=0.5]{{Ta=Tb_b=1.3_localSelection}.pdf}
    ~
    \includegraphics[scale=0.5]{{Ta=0.4_Tb=0.3-0.5_b=1.3_localSelection}.pdf} \\
  \caption{
  \textbf{Evolution of cooperation in a structured population with local selection.}
  The expected frequency of cooperators in a structured population after 10,000 generations is shown (red for 0\%, green for 100\%) as function of both the cost of cooperation ($c$) on the y-axis, and the symmetric horizontal transmission rate ($T=T_A=T_B$) on the x-axis of panel (a), or the transmission bias $T_A-T_B$ on the x-axis of panel (b).
  The population evolves on a 100-by-100 grid.
  Cooperation and horizontal transmission are both local between neighboring sites, and each site had 8 neighbors.
  Selection operates locally (see \autoref{fig:spatial} for results from a model with global selection).
  The black curves represent the cost thresholds for the evolution of cooperation in a well-mixed population with interaction-transmission association, where $\alpha=1/8$ in \autoref{eq:equal_transmission} for panel~(a) and in \autoref{eq:cost_thresholds_v_threshold} for panel (b).
  The population evolves on a 100-by-100 grid.
  Simulations were stopped at generation 10,000 or if one of the phenotypes fixed.
  50 simulations were executed for each parameter set.
  Here, benefit of cooperation, $b=1.3$; perfect vertical transmission $v=1$;
  \textbf{(a)} Symmetric horizontal transmission, $T=T_A=T_B$;
  \textbf{(b)} Horizontal transmission rates $T_A=0.4$, $0.3<T_B<0.5$;
  }
  \label{fig:spatial_local_selection}
\end{figure}

>>>>

% Abstract
\begin{abstract}
Cultural evolution of cooperation under vertical and non-vertical cultural transmission is studied, and conditions are found for fixation and coexistence of cooperation and defection. 
The evolution of cooperation is facilitated by its horizontal transmission and by an association between social interactions and horizontal transmission.
The effect of oblique transmission depends on the horizontal transmission bias.
Stable polymorphism of cooperation and defection can occur, and
when it does, reduced association between social interactions and horizontal transmission evolves, which leads to a decreased frequency of cooperation and lower population mean fitness.
The deterministic conditions are compared to outcomes of stochastic simulations of structured populations.
Parallels are drawn with Hamilton's rule incorporating assortment and effective relatedness.
\end{abstract}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%
Cooperation and defection can coexist at frequencies $\hat{p}^*$ and $1-\hat{p}^*$ (\autoref{eq:general_equilibrium}). 
When it is feasible, this equilibrium is stable or unstable under the conditions of Result \ref{result:vert_obli_hori}, parts 3 and 4, respectively. The yellow and blue areas in Figures~\ref{fig:equilibria_v1} and \ref{fig:equilibria} show cases of stable and unstable polymorphism, respectively.
When $\hat{p}^*$ is unstable, cooperation will fix if its initial frequency is $\hat{p}>\hat{p}^*$, and defection will fix if $\hat{p}<\hat{p}^*$.
$\hat{p}^*$ is unstable when there is horizontal transmission bias for cooperation, $T_A>T_B$, and the cost is intermediate, $\gamma_1 < c < \gamma_2$.
\autoref{fig:equilibria_v1}d shows $\hat{p}'-\hat{p}$ as a function of $\hat{p}$.

@@

\begin{figure}[p]
  \centering       
    \includegraphics[width=0.45\textwidth]{Result3_TA_TB.eps}~
    \includegraphics[width=0.45\textwidth]{Result3_TB_TA.eps}
    \includegraphics[width=0.45\textwidth]{Result3_alpha_Vs_v_TA_TB.eps}~
    \includegraphics[width=0.45\textwidth]{Result3_alpha_Vs_v_TB_TA.eps}
    \caption{\textbf{Evolution of cooperation under vertical, oblique, and horizontal cultural transmission.} 
    The figure shows parameter ranges for global fixation of cooperation (green), global fixation of defection (red), fixation of either cooperation or defection depending on the initial conditions, i.e. unstable polymorphism (blue), and stable polymorphism of cooperation and defection (yellow).
	In all cases the vertical transmission rate $v$ is on the x-axis.
	(\textbf{a-b})~Cost of cooperation $c$ is on the y-axis and the cost thresholds $\gamma_1$ and $\gamma_2$ (\autoref{eq:cost_thresholds_v_threshold}) are represented by the solid and dashed lines, respectively. 
    (\textbf{c-d})~interaction-transmission association $\alpha$ is on the y-axis and the interaction-transmission association thresholds $a_1$ and $a_2$ (Eqs.~\ref{eq:boundries_assortative_meeting_general_case}) are represented by the solid and dashed lines, respectively. 
    Horizontal transmission is biased in favor of cooperation, $T_A>T_B$, in (\textbf{a}) and (\textbf{c}), or defection, $T_A<T_B$, in (\textbf{b}) and (\textbf{d}).    
    Here, $T_A = 0.5$, and
    (\textbf{a}) $b=1.2$, , $T_B = 0.4$, $\alpha = 0.4$;
    (\textbf{b}) $b=2$, $T_B = 0.7$, $\alpha = 0.7$;
    (\textbf{c}) $b=1.2$, $T_B = 0.4$, $c=0.5$;
    (\textbf{d}) $b=2$, $T_B = 0.7$, $c=0.5$.
    }
    \label{fig:equilibria}
\end{figure}

@@

\begin{figure}[p]
  \centering       
    \includegraphics[width=0.45\textwidth]{Result2_c_zero_alpha.eps}
	  ~
    \includegraphics[width=0.45\textwidth]{Result2_c_non_zero_alpha.eps}
    ~
    \includegraphics[width=0.45\textwidth]{Result2_alpha.eps}   
    ~
    \includegraphics[width=0.48\textwidth]{coexistence_without_oblique.eps}
    \caption{\textbf{Evolution of cooperation under vertical and horizontal cultural transmission (v=1).} 
    The figure shows parameter ranges for global fixation of cooperation (green), global fixation of defection (red), fixation of either cooperation or defection depending on the initial conditions, i.e. unstable polymorphism (blue), and stable polymorphism of cooperation and defection (yellow).
    (\textbf{a-c})~The horizontal transmission bias ($T_A-T_B$) is on the x-axis.
    In panels (a) and (b), the cost of cooperation $c$ is on the y-axis and the cost thresholds $\gamma_1$ and $\gamma_2$ (\autoref{eq:cost_thresholds_v_threshold}) are the solid and dashed lines, respectively. 
	In panel (c), interaction-transmission association $\alpha$ is on the y-axis and the interaction-transmission association thresholds $a_1$ and $a_2$ (Eqs.~\ref{eq:boundries_assortative_meeting_general_case}) are the solid and dashed lines, respectively.
    Here, $b=1.3$, $T_A=0.4$, $v=1$, (a)~$\alpha = 0$, (b)~$\alpha = 0.7$, (c)~$c = 0.35$.    
    \textbf{(d)} Change in frequency of cooperation among juveniles ($\hat{p}'-\hat{p}$) as a function of the frequency ($\hat{p}$), see \autoref{eq:nextgen_juveniles}.
    The orange curve shows convergence to a stable polymorphism ($T_A = 0.4$, $T_B = 0.9$, $b = 12$, $c=0.35$, $v=1$, and $\alpha = 0.45$). % todo ref eq 12
    The blue curve shows fixation of either cooperation or defection, depending on the initial frequency ($T_A = 0.5$, $T_B = 0.1$, $b = 1.3$, $c=0.904$, $v=1$, and $\alpha = 0.4$).
    Markers show the three equilibria.
  	}
    \label{fig:equilibria_v1}
\end{figure}

@@

%%% Figure: Spatial model - global selection
\begin{figure}[p]
  \centering
	\includegraphics[width=0.3\textwidth]{{Ta=Tb_b=1.3_globalSelection_a}.pdf}
    ~
	\includegraphics[width=0.3\textwidth]{{Ta=0.4_Tb=0.3-0.5_b=1.3_globalSelection_withSmallBox_b}.pdf} 
    ~
    \includegraphics[width=0.36\textwidth]{{timeSereis_b=1.3_TA=0.4_TB=0.435_c=0.02_globalSelection_normalizeByNumInteractions_c}.pdf}
  \caption{
  \textbf{Evolution of cooperation in a structured population.}
	\textbf{(a-b)}~The expected frequency of cooperators in a structured population after 10,000 generations is shown (red for 0\%, green for 100\%) as a function of both the cost of cooperation, $c$, on the y-axis, and either the symmetric horizontal transmission rate, $T=T_A=T_B$, on the x-axis of panel~(a), or the transmission bias, $T_A-T_B$, on the x-axis of panel~(b).
  Black curves represent the cost thresholds for the evolution of cooperation in a well-mixed population with interaction-transmission association, where $\alpha=1/8$ in inequality~\ref{eq:equal_transmission} for  panel~(a) and in inequality~\ref{eq:cost_thresholds_v_threshold} for~(b).
	The inset in panel~(b) focuses on an area of the parameter range in which neither phenotype is fixed throughout the simulation, maintaining a stochastic locally stable polymorphism~\citep{Karlin1975}.
	This stochastic polymorphism is illustrated in panel (c), which shows the frequency of cooperators (green) and defectors (red) over time for the parameter set marked by an~\textit{x} in panel~(b).
  In all cases, the population evolves on a 100-by-100 grid.
  Cooperation and horizontal transmission are both local between neighboring sites, and each site has 8 neighbors.
  Selection operates globally (see \autoref{fig:spatial_local_selection} for results from a model with local selection).
  Simulations were stopped at generation 10,000 or if one of the phenotypes fixed. 50 simulations were executed for each parameter set.
  Benefit of cooperation, $b=1.3$; perfect vertical transmission $v=1$;
  \textbf{(a)} Symmetric horizontal transmission, $T=T_A=T_B$;
  \textbf{(b)} Horizontal transmission rates $T_A=0.4$, $0.3<T_B<0.5$;
  \textbf{(c)} Horizontal transmission rates $T_A = 0.4 < T_B = 0.435$ and cost of cooperation $c=0.02$.
  }
  \label{fig:spatial}
\end{figure}

@@

%%% Figure: Spatial model - local selection
\begin{figure}[h]
  \centering
   \includegraphics[scale=0.5]{{Ta=Tb_b=1.3_localSelection}.pdf}
    ~
    \includegraphics[scale=0.5]{{Ta=0.4_Tb=0.3-0.5_b=1.3_localSelection}.pdf} \\
  \caption{
  \textbf{Evolution of cooperation in a structured population with local selection.}
  The expected frequency of cooperators in a structured population after 10,000 generations is shown (red for 0\%, green for 100\%) as function of both the cost of cooperation ($c$) on the y-axis, and the symmetric horizontal transmission rate ($T=T_A=T_B$) on the x-axis of panel (a), or the transmission bias $T_A-T_B$ on the x-axis of panel (b).
  The population evolves on a 100-by-100 grid.
  Cooperation and horizontal transmission are both local between neighboring sites, and each site had 8 neighbors.
  Selection operates locally (see \autoref{fig:spatial} for results from a model with global selection).
  The black curves represent the cost thresholds for the evolution of cooperation in a well-mixed population with interaction-transmission association, where $\alpha=1/8$ in inequality~\ref{eq:equal_transmission} for panel~(a) and in Eqs.~\ref{eq:cost_thresholds_v_threshold} for panel (b).
  The population evolves on a 100-by-100 grid.
  Simulations were stopped at generation 10,000 or if one of the phenotypes fixed.
  50 simulations were executed for each parameter set.
  Here, benefit of cooperation, $b=1.3$; perfect vertical transmission $v=1$;
  \textbf{(a)} Symmetric horizontal transmission, $T=T_A=T_B$;
  \textbf{(b)} Horizontal transmission rates $T_A=0.4$, $0.3<T_B<0.5$;
  }
  \label{fig:spatial_local_selection}
\end{figure}
